---
title: "GSRD Evaluation Report"
author: "Han Zhang"
date: "`r Sys.Date()`"

output: 
  pdf_document: 
    latex_engine: xelatex
    extra_dependencies: 
      - seqsplit
      - makecell
      - threeparttable
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1.Load packages and dataset

```{r}
library(tidyverse)
library(readxl)
library(knitr)
library(kableExtra)
library(ggplot2)
library(gridExtra)
library(reshape2)
library(scales)
library(maps)

gsrd <- read.csv("data/GSRD_main.csv", sep = ",")
```

# 2.Basic information of dataset

```{r,results='hold'}
n_entries <- nrow(gsrd)
n_columns <- ncol(gsrd)
column_names <- colnames(gsrd)

column_types <- vapply(gsrd, function(x) paste(class(x), collapse = ", "), character(1))

column_table <- tibble(
  `Column Number` = seq_along(column_names),
  `Column Name`   = column_names,
  `Data Type`     = column_types
)

cat(paste(rep("-", 80), collapse = ""), "\n\n")  
cat("Number of entries:", n_entries, "\n")
cat("Number of columns:", n_columns, "\n\n")

kable(column_table, 
      caption = "List of columns in the dataset (with data types)",
      booktabs = TRUE) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE
  )

cat(paste(rep("-", 80), collapse = ""), "\n\n")

```

# 3.Detailed information check

## 3.1 Reference and Location

### 3.1.1 DATAID, REFID and REF

```{r, results='hold'}
# Calculate unique counts for reference-related columns
unique_dataid <- n_distinct(gsrd$DATAID)
unique_refid <- n_distinct(gsrd$REFID)
unique_ref <- n_distinct(gsrd$REF)

# Get all unique references (sorted alphabetically)
all_unique_refs <- gsrd %>%
  select(REF) %>%
  distinct() %>%
  drop_na() %>%
  arrange(REF) %>%
  pull(REF)

cat(paste(rep("-", 80), collapse = ""), "\n\n")  
# Print num of dataid and refid
cat("Number of DATAID:", unique_dataid, "\n")
cat("Number of REFID:", unique_refid, "\n")
# Print references in a formatted list
cat("References", "\n")
cat("1. ", all_unique_refs[1], "\n", sep = "")
for (i in 2:length(all_unique_refs)) {
  cat(i, ". ", all_unique_refs[i], "\n", sep = "")
}
cat(paste(rep("-", 80), collapse = ""), "\n\n")  
```

### 3.1.2 LAT,LON and ELE

```{r, results='hold'}

# check the range of latitude, longitude and elevation
site_continuous <- gsrd %>%
  select(LAT, LON) %>%
  summarise(
    var = colnames(.),
    min = map_dbl(., min, na.rm = TRUE),
    max = map_dbl(., max, na.rm = TRUE),
    na_fraction = map_dbl(., ~mean(is.na(.))*100) #fraction of NA number
  )

kable(site_continuous, caption = "site location") %>%
  kable_styling(bootstrap_options = "striped")

site_ele <- gsrd %>%
  select(ELE) %>%
  summarise(
    var = colnames(.),
    na_fraction = map_dbl(., ~mean(is.na(.))*100) #fraction of NA number
  )

kable(site_ele, caption = "site elevation") %>%
  kable_styling(bootstrap_options = "striped")

# figure: global distribution
ggplot(gsrd, aes(x = LON, y = LAT)) +
  borders("world", colour = "gray50") +
  geom_point(alpha = 0.5) +
  coord_quickmap(xlim = c(-180, 180), ylim = c(-90, 90)) +
  scale_x_continuous(breaks = seq(-180, 180, 60), expand = c(0, 0)) +
  scale_y_continuous(breaks = seq(-90, 90, 30),  expand = c(0, 0)) +
  theme_minimal() +
  labs(title = "Site distribution", x = "Longitude", y = "Latitude")

```

## 3.2 Climate

including CLIMATE_ZONE, MAT, MAT_ORIGIN, MAP, MAP_ORIGIN, Tg and Alpha

```{r, results='hold'}

# climate_zone
climate_counts <- gsrd %>%
  count(CLIMATE_ZONE, name = "num") %>%
  arrange(desc(num))

kable(climate_counts, caption = "site climatic distribution") %>%
  kable_styling(bootstrap_options = "striped")

#* statistical description of mean annual temperature (MAT), 
#* mean annual precipitation (MAP), the mean temperature of growing season(Tg),
#* and Alpha calculated by SPLASH
climate_stats <- gsrd %>%
  select(MAT, MAP, Tg, Alpha) %>%
  summarise(
   var = colnames(.),
    mean = map_dbl(., mean, na.rm = TRUE),
    std = map_dbl(., sd, na.rm = TRUE),
    min = map_dbl(., min, na.rm = TRUE),
    q25 = map_dbl(., ~quantile(., 0.25, na.rm = TRUE)),  # 25%
    median = map_dbl(., ~quantile(., 0.50, na.rm = TRUE)),  # 50%
    q75 = map_dbl(., ~quantile(., 0.75, na.rm = TRUE)),  # 75%
    max = map_dbl(., max, na.rm = TRUE),
    count = map_dbl(., ~sum(!is.na(.))),
    na_fraction = map_dbl(., ~mean(is.na(.))*100)
  )

# Origin of MAT and MAP（MAT_ORIGIN, MAP_ORIGIN）
origin_counts <- gsrd %>%
  count(MAT_ORIGIN, name = "MAT_origin") %>%
  full_join(
    gsrd %>% count(MAP_ORIGIN, name = "MAP_origin"),
    by = c("MAT_ORIGIN" = "MAP_ORIGIN")
  )%>%
  rename(source="MAT_ORIGIN")

# output
kable(climate_stats, digits = 2, caption = "climate statistical description") %>%
  kable_styling(bootstrap_options = "striped")

kable(origin_counts, caption = "climate source") %>%
  kable_styling(bootstrap_options = "striped")

```

## 3.3 Measurement

### 3.3.1 SEASONAL, YEAR, MONTH, DOY

```{r}

# Calculate counts for SEASONAL categories (Y/N)
seasonal_stats <- gsrd %>%
  group_by(SEASONAL) %>%
  summarise(Count = n(), .groups = "drop") %>%
  complete(SEASONAL = c("Y", "N"), fill = list(Count = 0)) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# Calculate unique values and data availability for temporal variables
temporal_metrics <- tibble(
  Variable = c("YEAR", "MONTH", "DOY"),
  Unique_Values = c(
    n_distinct(gsrd$YEAR, na.rm = TRUE),
    n_distinct(gsrd$MONTH, na.rm = TRUE),
    n_distinct(gsrd$DOY, na.rm = TRUE)
  ),
  Observations_with_Data = c(
    sum(!is.na(gsrd$YEAR)),
    sum(!is.na(gsrd$MONTH)),
    sum(!is.na(gsrd$DOY))
  ),
  Data_Availability = paste0(
    round(Observations_with_Data / nrow(gsrd) * 100, 1), "%"
  )
)


kable(seasonal_stats,
      col.names = c("SEASONAL Category", "Count", "Percentage (%)"),
      caption = "Distribution of SEASONAL measurements",
      booktabs = TRUE) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE)

temporal_metrics %>%
  kable(
    col.names = c("Variable", "Unique Values", "Observations with Data", "Data Availability"),
    caption = "Summary of temporal measurement variables",
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(4) %>%
  footnote(
    general = "Data Availability is calculated as the percentage of 
    non-missing values relative to total observations.",
    general_title = "Note:",
    escape=FALSE
  )


```

### 3.3.2 MEASUREMENT and MTEM

```{r}
# Calculate counts for MEASUREMENT categories (FIELD/LAB)
measurement_counts <- gsrd %>%
  filter(!is.na(MEASUREMENT)) %>%
  count(MEASUREMENT) %>%
  complete(MEASUREMENT = c("FIELD", "LAB"), fill = list(n = 0)) %>%  
  rename(Category = MEASUREMENT, Count = n) %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# Calculate descriptive statistics for MTEM (measurement temperature)
mtem_stats <- gsrd %>%
  filter(!is.na(MTEM)) %>%
  summarise(
    Variable = "MTEM (Measurement Temperature)",
    N = n(),
    Missing = nrow(gsrd) - n(),
    Mean = round(mean(MTEM), 1),
    SD = round(sd(MTEM), 1),
    Min = round(min(MTEM), 1),
    Max = round(max(MTEM), 1),
    Median = round(median(MTEM), 1),
    `25th Percentile` = round(quantile(MTEM, 0.25), 1),
    `75th Percentile` = round(quantile(MTEM, 0.75), 1)
  )

measurement_counts %>%
  kable(
    caption = "Distribution of measurement methods",
    booktabs = TRUE,
    col.names = c("Measurement Method", "Count", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "3cm")


mtem_long <- mtem_stats %>%
  pivot_longer(cols = -Variable, names_to = "Statistic", values_to = "Value")

mtem_long %>%
  dplyr::select(Statistic, Value) %>%
  kable(
    caption = "Descriptive statistics for measurement temperature (°C)",
    booktabs = TRUE,
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(2, width = "2cm") %>%
  pack_rows(
    index = setNames(nrow(mtem_long), unique(mtem_long$Variable))  
  ) %>%
  footnote(
    general = "N = number of non-missing observations; SD = standard deviation",
    general_title = "Note:"
  )



```

## 3.4 Plant trait

### 3.4.1 ORIGINAL_SPNAME, FINAL_SPNAME, AUTHORITY, GENUS, FAMILY, GROUP and PFT

```{r}
invisible(gc())
# Calculate unique counts for taxonomic variables
taxonomic_unique <- tibble(
  Variable = c("ORIGINAL_SPNAME", "FINAL_SPNAME", "AUTHORITY", 
               "GENUS", "FAMILY", "GROUP", "PFT"),
  Unique_Values = c(
    n_distinct(gsrd$ORIGINAL_SPNAME, na.rm = TRUE),
    n_distinct(gsrd$FINAL_SPNAME, na.rm = TRUE),
    n_distinct(gsrd$AUTHORITY, na.rm = TRUE),
    n_distinct(gsrd$GENUS, na.rm = TRUE),
    n_distinct(gsrd$FAMILY, na.rm = TRUE),
    n_distinct(gsrd$GROUP, na.rm = TRUE),
    n_distinct(gsrd$PFT, na.rm = TRUE)
  )
)

taxonomic_unique %>%
  kable(
    caption = "Unique value counts for taxonomic and functional variables",
    booktabs = TRUE,
    col.names = c("Variable", "Number of Unique Values")
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  footnote(
    general = "PFT = Plant Functional Type; GROUP = Gymnosperm or Angiosperm",
    general_title = "Note:"
  )

rm(taxonomic_unique) 
invisible(gc())

```


```{r}
# Data preparation: process taxonomic variables and handle NAs
gsrd_taxonomy <- gsrd %>%
  select(FINAL_SPNAME, AUTHORITY, GENUS, FAMILY, GROUP, PFT) %>%
  mutate(
    # Convert all to character to avoid type issues
    across(everything(), ~as.character(.)),
    # Combine species name and authority, handling NAs
    Species_Authority = case_when(
      !is.na(FINAL_SPNAME) & !is.na(AUTHORITY) ~ paste(FINAL_SPNAME, 
                                                       "(", AUTHORITY, ")", 
                                                       sep = ""),
      !is.na(FINAL_SPNAME) & is.na(AUTHORITY) ~ FINAL_SPNAME,
      TRUE ~ NA_character_
    )
  )

# 1. Unique species (FINAL_SPNAME + AUTHORITY) as formatted text list
unique_species <- gsrd_taxonomy %>%
  filter(!is.na(Species_Authority)) %>%
  distinct(Species_Authority) %>%
  arrange(Species_Authority) %>%  # Sort alphabetically for readability
  pull(Species_Authority)

## Unique species list
### Species List (FINAL_SPNAME + AUTHORITY)
cat(paste("-", unique_species, collapse = "\n"))
cat(paste(rep("-", 80), collapse = ""), "\n\n")  

# 2. Unique genera as comma-separated text
unique_genera <- gsrd_taxonomy %>%
  filter(!is.na(GENUS)) %>%
  distinct(GENUS) %>%
  arrange(GENUS) %>%
  pull(GENUS) %>%
  paste(collapse = ", ")

data.frame(Genera_List = unique_genera) %>%
  kable(
    caption = "Unique genera list",
    booktabs = TRUE,
    col.names = c("Genera List")
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, width = "15cm") %>% 
  row_spec(0, bold = TRUE)

# 3. Unique families as comma-separated text
unique_families <- gsrd_taxonomy %>%
  filter(!is.na(FAMILY)) %>%
  distinct(FAMILY) %>%
  arrange(FAMILY) %>%
  pull(FAMILY) %>%
  paste(collapse = ", ")

data.frame(Families_List = unique_families) %>%
  kable(
    caption = "Unique families list",
    booktabs = TRUE,
    col.names = c("Families List")
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, width = "15cm") %>%  
  row_spec(0, bold = TRUE)

# 4. Distribution of GROUP (Gymnosperm/Angiosperm)
group_dist <- gsrd_taxonomy %>%
  filter(!is.na(GROUP)) %>%
  count(GROUP, name = "Count") %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1)) %>%
  rename("Group" = GROUP)

# 5. Distribution of PFT (Plant Functional Types)
pft_dist <- gsrd_taxonomy %>%
  filter(!is.na(PFT)) %>%
  count(PFT, name = "Count") %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1)) %>%
  rename("Plant Functional Type (PFT)" = PFT) %>%
  arrange(desc(Count))  # Sort by abundance


group_dist %>%
  kable(
    caption = "Distribution of observations across plant groups",
    booktabs = TRUE,
    col.names = c("Group", "Number of Observations", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(3)

pft_dist %>%
  kable(
    caption = "Distribution of observations across plant functional types",
    booktabs = TRUE,
    col.names = c("Plant Functional Type (PFT)", "Number of Observations", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(3)


```

### 3.4.2 AGE, DIAMETER, WD and RAP

```{r}

# Data preparation: convert to numeric, handle non-numeric values, 
# and ensure AGE is integer
gsrd_traits <- gsrd %>%
  select(AGE, DIAMETER, WD, RAP) %>%
  mutate(
    across(everything(), ~as.numeric(as.character(.))),  # Ensure numeric type
    AGE = ifelse(is.na(AGE), NA, round(AGE)),  # Round AGE to nearest integer and preserve NA
    RAP = factor(RAP)                 # Treat RAP as categorical
  )

# 1. Unique value distribution of AGE (sorted numerically)
age_dist <- gsrd_traits %>%
  filter(!is.na(AGE)) %>%
  count(AGE, name = "Count") %>%
  arrange(AGE) %>%  # Sort by age value
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# 2. Diameter distribution (binned into intervals)
diameter_bins <- gsrd_traits %>%
  filter(!is.na(DIAMETER)) %>%
  mutate(
    Diameter_Bin = cut(
      DIAMETER,
      breaks = c(0, 10, 20, 30, 50, Inf),  # Bins: 0-10, 10-20, ..., >50 cm
      labels = c("0-10 cm", "10-20 cm", "20-30 cm", "30-50 cm", ">50 cm"),
      include.lowest = TRUE
    )
  ) %>%
  count(Diameter_Bin, name = "Count") %>%
  mutate(Percentage = round(Count / sum(Count) * 100, 1))

# 3. Descriptive statistics for WD (wood density)
wd_stats <- gsrd_traits %>%
  filter(!is.na(WD)) %>%
  summarise(
    Variable = "WD (g cm⁻³)",
    N = n(),
    Mean = round(mean(WD), 3),
    SD = round(sd(WD), 3),
    Min = round(min(WD), 3),
    Max = round(max(WD), 3),
    Median = round(median(WD), 3),
    `25th Percentile` = round(quantile(WD, 0.25), 3),
    `75th Percentile` = round(quantile(WD, 0.75), 3)
  )

# 4. Count distribution for RAP (two categorical values)
rap_dist <- gsrd_traits %>%
  filter(!is.na(RAP)) %>%
  count(RAP, name = "Count") %>%
  mutate(
    RAP = as.character(RAP),
    Percentage = round(Count / sum(Count) * 100, 1)
  )


age_dist %>%
  kable(
    caption = "Unique values and counts for AGE (stand age in years)",
    booktabs = TRUE,
    col.names = c("AGE (years)", "Count", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE)

diameter_bins %>%
  kable(
    caption = "Distribution of DIAMETER (cm) across size intervals",
    booktabs = TRUE,
    col.names = c("Diameter Interval", "Count", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE)


wd_long <- wd_stats %>%
  tidyr::pivot_longer(cols = -Variable, names_to = "Statistic", values_to = "Value")

wd_long %>%
  dplyr::select(Statistic, Value) %>%
  kableExtra::kable(
    caption = "Descriptive statistics for WD (wood density)",
    booktabs = TRUE,
    longtable = TRUE,                     
    col.names = c("Statistic", "Value")
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("striped", "repeat_header", "scale_down"),  
    full_width = FALSE,
    position = "center",
    font_size = 8
  ) %>%
  kableExtra::pack_rows(
    index = setNames(nrow(wd_long), "WD (g cm$^{-3}$)")  
  ) %>%
  kableExtra::footnote(
    general = "WD = Wood Density",
    general_title = "Note:"
  )



rap_dist %>%
  kable(
    caption = "Distribution of RAP (parenchyma tissue fraction)",
    booktabs = TRUE,
    col.names = c("RAP Value", "Count", "Percentage (%)")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  footnote(
  general = paste0(
      "RAP values correspond to angiosperms (24.6%) and gymnosperms (7.6%) ."
  ),
  general_title = "Note:"
)

```

## 3.5 Stem Respiration Rate

### 3.5.1 Original stem respiration rate data
RS_AREA, RS_VOL, RS_MASS

```{r}
gsrd_resp <- gsrd %>%
  select(RS_ORIGIN, OriginUnit, RS_AREA, RS_VOL, RS_MASS) %>%
  mutate(
    RS_ORIGIN = as.character(RS_ORIGIN),
    OriginUnit = as.character(OriginUnit)
  )

# 1. RS_ORIGIN and OriginUnit distribution
rs_origin_summary <- gsrd_resp %>%
  filter(!is.na(RS_ORIGIN)) %>%
  group_by(RS_ORIGIN, OriginUnit) %>%
  summarise(Count = n(), .groups = "drop")

# 2. Descriptive stats for respiration variables (simplified)
resp_stats <- gsrd_resp %>%
  select(RS_AREA, RS_VOL, RS_MASS) %>%
  summarise(
    Variable = names(.),
    N = colSums(!is.na(.)),
    Mean = round(colMeans(., na.rm = TRUE), 2),
    SD = round(sapply(., sd, na.rm = TRUE), 2),
    Min = round(sapply(., min, na.rm = TRUE), 2),
    Max = round(sapply(., max, na.rm = TRUE), 2),
    Median = round(sapply(., median, na.rm = TRUE), 2),
    `25th Percentile` = round(sapply(., function(x) quantile(x, 0.25, na.rm = TRUE)), 2),
    `75th Percentile` = round(sapply(., function(x) quantile(x, 0.75, na.rm = TRUE)), 2)
  )
rs_origin_summary %>%
  kable(
    booktabs = TRUE,
    caption = "Summary of RS\\_ORIGIN categories and original units.",
    col.names = c("RS_ORIGIN Category", "OriginUnit", "Number of Observations")
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped"), 
                full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)

resp_stats <- gsrd_resp %>%
  select(RS_AREA, RS_VOL, RS_MASS) %>%
  summarise(
    Variable = names(.),
    N = colSums(!is.na(.)),
    Mean = round(colMeans(., na.rm = TRUE), 2),
    SD = round(sapply(., sd, na.rm = TRUE), 2),
    Min = sapply(., function(x) {
      min_val <- min(x, na.rm = TRUE)
      if (abs(min_val) < 0.01 && min_val != 0) {
        formatC(min_val, format = "e", digits = 2)
      } else {
        as.character(round(min_val, 2))  
      }
    }),
    Max = round(sapply(., max, na.rm = TRUE), 2),
    Median = round(sapply(., median, na.rm = TRUE), 2),
    `25th Percentile` = round(sapply(., function(x) quantile(x, 0.25, na.rm = TRUE)), 2),
    `75th Percentile` = round(sapply(., function(x) quantile(x, 0.75, na.rm = TRUE)), 2)
  )

resp_stats %>%
  kable(
    booktabs = TRUE,
    caption = "Descriptive statistics for RS\\_AREA, RS\\_VOL, and RS\\_MASS."
  ) %>%
  kable_styling(latex_options = c("hold_position", "striped"), 
                full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE)


```

### 3.5.2 Standardization stem respiration rate data
RSMASS_CAL, RS_MASS25, LNRS25, LNRS25_RAP

```{r}
# Function to calculate and format statistics for a single variable
get_stats <- function(data, var_name) {
  var_data <- pull(data, {{var_name}})
  tibble(
    Statistic = c(
      "Number of non-missing observations",
      "Mean", "Standard deviation (SD)",
      "Minimum", "Maximum", "Median",
      "25th percentile", "75th percentile"
    ),
    Value = c(
      sum(!is.na(var_data)),
      round(mean(var_data, na.rm = TRUE), 3),
      round(sd(var_data, na.rm = TRUE), 3),
      round(min(var_data, na.rm = TRUE), 3),
      round(max(var_data, na.rm = TRUE), 3),
      round(median(var_data, na.rm = TRUE), 3),
      round(quantile(var_data, 0.25, na.rm = TRUE), 3),
      round(quantile(var_data, 0.75, na.rm = TRUE), 3)
    )
  )
}

# Calculate statistics for each variable
stats_rsmass_cal <- get_stats(gsrd, RSMASS_CAL)
stats_rs_mass25 <- get_stats(gsrd, RS_MASS25)
stats_lnrs25 <- get_stats(gsrd, LNRS25)
stats_lnrs25_rap <- get_stats(gsrd, LNRS25_RAP)

stats_rsmass_cal %>%
  kable(
    caption = "Descriptive statistics for RSMASS\\_CAL",
    booktabs = TRUE,
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm")%>%
  footnote(
    general = 
    "RSMASS_CAL is derived from standardized calculations of raw respiration data.",
    general_title = "Note:",
    escape = TRUE
  )


stats_rs_mass25 %>%
  kable(
    caption = "Descriptive statistics for RS\\_MASS25",
    booktabs = TRUE,
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm")%>%
  footnote(
    general = "RS_MASS25 is temperature standardized mass-based respiration data.",
    general_title = "Note:"
  )


stats_lnrs25 %>%
  kable(
    caption = "Descriptive statistics for LNRS25",
    booktabs = TRUE,
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm")%>%
  footnote(
    general = "LNRS25 is natural-logged RSMASS25.",
    general_title = "Note:"
  )


stats_lnrs25_rap %>%
  kable(
    caption = "Descriptive statistics for LNRS25\\_RAP",
    booktabs = TRUE,
    col.names = c("Statistic", "Value")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, width = "5cm") %>%
  column_spec(2, width = "3cm")%>%
  footnote(
    general = 
    "LNRS25_RAP is natural-logged RSMASS25 considering parenchyma tissue fractions.",
    general_title = "Note:",
    escape = TRUE
  )

```

# 4. Technical validation
## 4.1 Latitude and longitude validation
```{r}
# valid ranges: Lat -90 to 90; Lon -180 to 180
coord_check <- gsrd %>%
  mutate(
    LAT_valid = ifelse(LAT >= -90 & LAT <= 90 & !is.na(LAT), "Valid", "Invalid"),
    LON_valid = ifelse(LON >= -180 & LON <= 180 & !is.na(LON), "Valid", "Invalid")
  ) %>%
  count(LAT_valid, LON_valid) %>%
  rename(`Latitude Status` = LAT_valid, `Longitude Status` = LON_valid, `Count` = n)

coord_check %>%
  kable(
    caption = "Validation results for latitude and longitude",
    booktabs = TRUE
  ) %>%
  kable_styling(
    latex_options = "hold_position",
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE)

```

## 4.2 Main sources of potential outliers under global vs. site-level IQR
```{r}
## =========================
## Parameters
## =========================
digits_site <- 3   # Site definition: round(LAT/LON, digits_site)
min_n_site  <- 10  # Minimum n within a site to compute site-level IQR; 
                   #otherwise fallback to global thresholds
top_n_pairs <- 30  # Top N REFID×SITE_ID groups to display in each table

## =========================
## 0) Basic checks
## =========================
required_cols <- c("REFID", "LAT", "LON", "RS_MASS25")
stopifnot(all(required_cols %in% names(gsrd)))

## =========================
## 1) Create SITE_ID (rounded LAT/LON)
## =========================
gsrd2 <- gsrd %>%
  mutate(
    LAT_r = round(LAT, digits_site),
    LON_r = round(LON, digits_site),
    SITE_ID = if_else(
      is.na(LAT_r) | is.na(LON_r),
      NA_character_,
      paste0(
        "lat", formatC(LAT_r, format = "f", digits = digits_site),
        "_lon", formatC(LON_r, format = "f", digits = digits_site)
      )
    )
  )

## =========================
## 2) Compute global 3×IQR thresholds (once)
## =========================
x <- gsrd2$RS_MASS25
iqr_g <- IQR(x, na.rm = TRUE)
q1_g  <- quantile(x, 0.25, na.rm = TRUE)
q3_g  <- quantile(x, 0.75, na.rm = TRUE)

low_g  <- max(0, as.numeric(q1_g - 3 * iqr_g))
high_g <- as.numeric(q3_g + 3 * iqr_g)

## =========================
## 3) Global outlier flag (global 3×IQR)
## =========================
gsrd2 <- gsrd2 %>%
  mutate(
    outlier_global = case_when(
      is.na(RS_MASS25) ~ NA,
      TRUE ~ (RS_MASS25 < low_g | RS_MASS25 > high_g)
    )
  )

## =========================
## 4) Site-level thresholds (site 3×IQR; small-n fallback to global)
## =========================
site_thr <- gsrd2 %>%
  filter(!is.na(SITE_ID)) %>%
  group_by(SITE_ID) %>%
  summarise(
    n_site = sum(!is.na(RS_MASS25)),
    q1_s   = if (n_site >= min_n_site) quantile(RS_MASS25, 0.25, na.rm = TRUE) 
    else NA_real_,
    q3_s   = if (n_site >= min_n_site) quantile(RS_MASS25, 0.75, na.rm = TRUE) 
    else NA_real_,
    iqr_s  = if (n_site >= min_n_site) IQR(RS_MASS25, na.rm = TRUE) 
    else NA_real_,
    .groups = "drop"
  ) %>%
  mutate(
    low_s  = q1_s - 3 * iqr_s,
    high_s = q3_s + 3 * iqr_s
  )

gsrd2 <- gsrd2 %>%
  left_join(site_thr, by = "SITE_ID") %>%
  mutate(
    thr_low_siteiqr = case_when(
      is.na(RS_MASS25) ~ NA_real_,
      !is.na(n_site) & n_site >= min_n_site ~ low_s,
      TRUE ~ low_g
    ),
    thr_high_siteiqr = case_when(
      is.na(RS_MASS25) ~ NA_real_,
      !is.na(n_site) & n_site >= min_n_site ~ high_s,
      TRUE ~ high_g
    ),
    thr_source_siteiqr = case_when(
      is.na(RS_MASS25) ~ NA_character_,
      !is.na(n_site) & n_site >= min_n_site ~ paste0("site_3xIQR(SITE_ID_round", 
                                                     digits_site, ")"),
      TRUE ~ "global_3xIQR_fallback"
    ),
    outlier_siteiqr = case_when(
      is.na(RS_MASS25) ~ NA,
      TRUE ~ (RS_MASS25 < thr_low_siteiqr | RS_MASS25 > thr_high_siteiqr)
    )
  )

## =========================
## 5) Table 1: Global IQR (REFID×SITE_ID outlier distribution)
## =========================
tbl_global_ref_site <- gsrd2 %>%
  filter(!is.na(REFID), !is.na(SITE_ID), !is.na(outlier_global)) %>%
  group_by(REFID, SITE_ID) %>%
  summarise(
    n_nonmiss = n(),
    outlier_n = sum(outlier_global),
    outlier_pct = round(100 * mean(outlier_global), 2),
    LAT = first(LAT_r),
    LON = first(LON_r),
    thr_low  = round(low_g, 3),
    thr_high = round(high_g, 3),
    .groups = "drop"
  ) %>%
  filter(outlier_n > 0) %>% 
  arrange(desc(outlier_n), desc(outlier_pct), desc(n_nonmiss)) %>%
  slice_head(n = top_n_pairs)

tbl_global_ref_site %>%
  dplyr::mutate(
    SITE_ID = kableExtra::linebreak(SITE_ID, align = "l")
  ) %>%
  dplyr::select(REFID, SITE_ID, n_nonmiss, outlier_n, outlier_pct, LAT, LON) %>%
  kable(
    caption = paste0(
      "Main sources of potential outliers under GLOBAL $3\\times IQR$ 
      (RS\\_MASS25; SITE\\_ID from rounded LAT/LON, digits=",
      digits_site, ")"
    ),
    booktabs = TRUE,
    align = "c",
    escape = FALSE,
    col.names = c("REFID", "SITE_ID", "n_nonmiss", 
                  "outlier_n", "outlier_pct", "LAT", "LON")
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped", "scale_down"),
    full_width = FALSE, position = "center",
    font_size = 8
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%
  column_spec(1, width = "1.8cm") %>%
  column_spec(2, width = "3.0cm") %>%
  footnote(
    general = kableExtra::linebreak(
      paste0(
        "Global thresholds are constant across rows: [", round(low_g, 3), ", ", 
        round(high_g, 3), "]. ",
        "Sorted by outlier count, then outlier proportion."
      ),
      align = "l"
    ),
    general_title = "Note:",
    escape = TRUE
  )


## =========================
## 6) Table 2: Site-level IQR (REFID×SITE_ID outlier distribution)
## =========================
tbl_siteiqr_ref_site <- gsrd2 %>%
  filter(!is.na(REFID), !is.na(SITE_ID), !is.na(outlier_siteiqr)) %>%
  group_by(REFID, SITE_ID) %>%
  summarise(
    n_nonmiss = n(),
    outlier_n = sum(outlier_siteiqr),
    outlier_pct = round(100 * mean(outlier_siteiqr), 2),
    LAT = first(LAT_r),
    LON = first(LON_r),
    site_n = first(n_site),
    thr_source = first(thr_source_siteiqr),
    thr_low  = round(first(thr_low_siteiqr), 3),
    thr_high = round(first(thr_high_siteiqr), 3),
    .groups = "drop"
  ) %>%
  filter(outlier_n > 0) %>% 
  arrange(desc(outlier_n), desc(outlier_pct), desc(n_nonmiss)) %>%
  slice_head(n = top_n_pairs)

tbl_siteiqr_ref_site %>%
  dplyr::transmute(
    REFID,
    SITE_ID = kableExtra::linebreak(SITE_ID, align = "l"),
    N = n_nonmiss,
    OutN = outlier_n,
    OutPct = outlier_pct,
    Thr = thr_source,
    Low = thr_low,
    High = thr_high
  ) %>%
  kable(
    caption = paste0(
      "Main sources of potential outliers under SITE-level 3xIQR (RS\\_MASS25; digits=",
      digits_site, ", min\\_n\\_site=", min_n_site, ")"
    ),
    booktabs = TRUE,
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("striped", "scale_down"), 
    full_width = FALSE, position = "center",
    font_size = 7
  ) %>%
  column_spec(2, width = "3.2cm") %>%
  footnote(
    general = kableExtra::linebreak(
      paste0(
        "Site thresholds computed within each SITE_ID when n_site >= ", min_n_site,
        "; otherwise global fallback [", round(low_g, 3), ", ", round(high_g, 3), "]."
      ),
      align = "l"
    ),
    general_title = "Note:",
    escape = TRUE
  )


```

####  Overall comparison: global vs. site-level IQR outlier rates

```{r}
outlier_summary <- gsrd2 %>%
  summarise(
    total_nonmiss      = sum(!is.na(RS_MASS25)),
    global_outlier_n   = sum(outlier_global, na.rm = TRUE),
    global_outlier_pct = round(100 * mean(outlier_global, na.rm = TRUE), 2),
    site_outlier_n     = sum(outlier_siteiqr, na.rm = TRUE),
    site_outlier_pct   = round(100 * mean(outlier_siteiqr, na.rm = TRUE), 2)
  )

outlier_summary %>%
  kable(
    caption = "Overall comparison of global vs. site-level outlier rates (3xIQR)",
    booktabs = TRUE,
    col.names = c("N", "Global N", "Global %", "Site N", "Site %"),
    align = "c"
  ) %>%
  kable_styling(
    latex_options = c("striped", "scale_down"), 
    full_width = FALSE,
    position = "center",
    font_size = 8
  ) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0")



```

## 4.3 IQR figures of stem respiration rates
Frequency plots and outlier detection (defined by IQR)
Violin plots + jittered points + red dashed lines for outlier thresholds

```{r, results='asis', echo=FALSE, fig.show='hold'}
# Keep only the three required variables
vars <- c("RS_AREA", "RS_VOL", "RS_MASS")
# Check if variables exist in the dataset
vars <- vars[vars %in% names(gsrd)]
if (length(vars) == 0) stop("None of the specified variables were found in 'gsrd'.")

# Replace Inf / -Inf values with NA
for (v in vars) gsrd[[v]][is.infinite(gsrd[[v]])] <- NA

# Function to calculate outlier thresholds (IQR method, 3*IQR)
outlier_thresholds <- function(x, iqr_mult = 3) {
  x <- x[!is.na(x)]
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - iqr_mult * iqr
  upper <- q3 + iqr_mult * iqr
  c(lower = as.numeric(lower), upper = as.numeric(upper))
}

# Function to plot a single-variable violin plot with outlier thresholds
plot_violin_with_outliers <- function(data, var, iqr_mult = 3) {
  x <- data[[var]]
  x_clean <- x[!is.na(x) & is.finite(x)]
  
  # Skip variables with insufficient numeric data
  if (length(x_clean) < 2) {
    return(ggplot() + ggtitle(paste(var, "- no numeric data")) + theme_void())
  }
  
  # Calculate outlier thresholds and summary statistics
  th <- outlier_thresholds(x_clean, iqr_mult)
  n_out <- sum(x_clean < th["lower"] | x_clean > th["upper"])
  total <- length(x_clean)
  pct <- round(100 * n_out / total, 2)
  
  # Create violin plot
  ggplot(data, aes(x = "", y = .data[[var]])) +
    geom_violin(fill = "grey80", color = "black") +
    geom_jitter(width = 0.15, size = 1, alpha = 0.6) +
    geom_hline(yintercept = th, color = "red", size = 0.8, linetype = "dashed") +
    theme_bw() +
    labs(
      title = paste0(var, " (outliers = ", n_out, "/", total, " = ", pct, "%)"),
      x = NULL, y = var
    ) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank()
    )
}

# Generate and output individual plots for each variable
for (v in vars) {
  # Render and output the individual plot
  print(plot_violin_with_outliers(gsrd, v, iqr_mult = 3))
  # Add page breaks between plots (for PDF output, optional)
  if (v != tail(vars, 1)) {
    if (knitr::is_latex_output()) {
      cat("\\newpage\n")  # Page break for PDF
    } else {
      cat("\n\n***\n\n")  # Separator for other formats
    }
  }
}
```

